Visualizing Contingency Tables

### 2006-04-22 trunk/vcd/paper/hsc.Rnw
more fixes
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@523 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
%FIXME% right margin labels get clipped
<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
          legend = FALSE, margins = c(3,1,1,3), panel_args = list(type = "assoc"))
@
\caption{Conditional mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(tab, panel = cotab_coindep, panel_args = list(margin = c(2.5, 0, 0.5, 0)))
@
\caption{Conditional mosaic plot for the \data{punishment} data.}
\label{fig:punishcotab}
\end{center}
\end{figure}

@@

\bibliographystyle{plain}
\bibliography{vcd}

>>>>

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a rectangular mosaic. The tiles are
obtained by recursive partitioning splits of a rectangle.
In the following, we describe the main idea of mosaic plots, 
Chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Figure~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Figure~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to use filled tiles to highlight information
of interest. Consider again the hospital example: in Figure~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing relationships among
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in Chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<doubleucb,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
doubledecker(aperm(aperm(UCBAdmissions, c(1,3,2))[2:1,,], c(2,3,1)), margins = c(left = 0, right = 5),col = rev(grey.colors(2)))
@
\caption{Doubledecker plot for the \data{UCBAdmissions} data.}
\label{fig:doubleucb}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbmosaic,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
mosaic(aperm(UCBAdmissions, c(3,2,1)), data = UCBAdmissions, split = TRUE,
shade = TRUE, keep = FALSE, residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE)) 
@
\caption{Mosaic plot for the \data{UCB admissions} data.}
\label{fig:ucbmosaic}
\end{center}
\end{figure}

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<ucbassoc,fig=TRUE,echo=FALSE,height=6,width=12,results=hide>>=
assoc(aperm(UCBAdmissions, c(3,2,1)), split = c(TRUE,FALSE,TRUE), shade = TRUE, 
residuals_type = "Pearson", residuals = ucb_max$residuals,
gp_args = list(p.value = ucb_max$p.value), rep = c(Admit = FALSE))
@
\caption{Association plot for the \data{UCB admissions} data.}
\label{fig:ucbassoc}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<UCBAdmissions,fig=TRUE,echo=FALSE,height=6,width=10,results=hide>>=
cotabplot(aperm(UCBAdmissions, c(2,1,3)), panel = cotab_coindep, shade = TRUE, 
          legend = FALSE, 
          panel_args = list(type = "assoc", margins = c(2,1,1,2), varnames = FALSE))
@
\caption{Conditional association plot for the \data{UCB admissions}
  data. For each individual association plot, gender is in the rows
  and admit is in the columns.}
\label{fig:ucbadmissions}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=\textwidth}
\begin{figure}
\begin{center}
<<punishcotab,fig=TRUE,echo=FALSE,height=10,width=10,results=hide>>=
cotabplot(tab, panel = cotab_coindep, panel_args = list(varnames = FALSE, 
margins = c(2, 1, 1, 2)))
@
\caption{Conditional mosaic plot for the \data{punishment} data. For
  each mosaic plot, memory is in the rows (first split) and attitude
  in the columns (second split).}
\label{fig:punishcotab}
\end{center}
\end{figure}

@@

\bibliographystyle{abbrv}
\bibliography{vcd}
